---
title: "Biological invasion increases carbon sequestration rates but not carbon stocks"
author:
    - name: Murphy, S
      url: https://scholar.google.com/citations?hl=en&user=jDGq9I4AAAAJ
      orcid: 0000-0002-1792-0351
      email: seamusrobertmurphy@gmail.com
date: today
abstract: >
 Whether biological invasion increases ecosystem carbon storage is contested, and the disagreement has consequences for how invaded land is valued under carbon policy. We show that the contest is largely artefactual, arising because two different quantities are both reported as sequestration. Invasion raises the *rate* at which carbon moves through an ecosystem; it does not reliably raise the *stock* held there. These are the two methods the IPCC treats as equally valid for land carbon accounting, gain-loss and stock-difference, and on invaded land they disagree. We fit parallel meta-analyses of both quantities: 425 paired invaded and uninvaded observations from 57 studies across eight soil carbon fractions for stocks, and 688 paired observations from 110 sites in 25 countries for fluxes and related processes. Soil organic carbon was +20.4% higher under invasion but the interval spanned zero and the estimate fell to +6.8% under equal weighting. Among observations reporting carbon as an areal stock rather than a concentration the effect was -0.8%, 95% CI -15.7 to +16.7, and no contributing study reported the bulk density needed to convert between the two. Process rates moved in the opposite direction: soil respiration rose +24.3%, carbon-acquiring enzyme activity +24.6% and microbial biomass +12.6%, while soil fauna declined 32.3%. The gap between the flux and areal stock responses was +25.3 percentage points, in the predicted direction but not significant (*p* = 0.156). The strongest support is internal to the stock data, where the invasion effect declined with the turnover time of the pool, being largest in the fastest-cycling fractions; mineral-associated carbon was too sparsely measured to resolve. Carbon crediting rewards evidence of rate while paying for durable stock. Where the two diverge, as they do under invasion, that mismatch is a systematic source of over-crediting and a route to bio-perversity.
keywords: 
    - Invasive species
    - Carbon sequestration 
    - Meta-analysis 
    - Gain-loss method 
    - Stock-difference method 
    - Forest carbon 
    - Soil organic carbon 

format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    page-layout: full
    toc: true
    number-sections: false
    embed-resources: true
    toc-location: right
    theme: cosmo
    code-fold: false
    code-tools: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: false
  comment: NA
engine: knitr
citeproc: true
bibliography: ../04.references/references.bib
csl: ../04.references/apa.csl
---

```{r}
#| label: setup
#| echo: true
#| output: false

# This manuscript is the analysis. Every model, table and figure below is computed when the
# document is rendered, from the two deposited datasets in 02.inputs/, and every number in the
# prose is an inline expression, so the text cannot disagree with the code that produced it.
# Nothing is fetched over the network. Base R throughout, with metafor for the models, readxl
# to open the stock deposit's workbook, and knitr for the tables.
suppressPackageStartupMessages({library(metafor); library(readxl)})
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(20260730)

# All paths below are relative to the repository root, so root.dir is set once here: knitr
# restores the document directory after every chunk, and setwd alone would not survive.
repo <- normalizePath("..")
knitr::opts_knit$set(root.dir = repo)
setwd(repo)
FIG <- file.path(repo, "03.outputs", "figures"); TBL <- file.path(repo, "03.outputs", "tables")
dir.create(FIG, recursive = TRUE, showWarnings = FALSE)
dir.create(TBL, recursive = TRUE, showWarnings = FALSE)

# ---- formatting helpers, used by the inline expressions throughout the text -------------
pct <- function(b) (exp(b) - 1) * 100
f1  <- function(x) sprintf("%.1f", x)
sg  <- function(x) sprintf("%+.1f", x)
f0  <- function(x) formatC(round(x), format = "d", big.mark = ",")
pf  <- function(p) if (p < 0.001) "< 0.001" else sprintf("= %.3f", p)
ci  <- function(f) sprintf("%s%%, 95%% CI %s to %s", sg(f$est), sg(f$lo), sg(f$hi))
cif <- function(f) sprintf("%s%%, 95%% CI %s to %s", sg(f$pct), sg(f$lo), sg(f$hi))

# ---- the model used throughout -----------------------------------------------------------
# Multilevel random effects with observations nested in the clustering unit, fitted by REML
# (Viechtbauer 2010), which accounts for the non-independence of several observations drawn
# from one study or one site (Noble et al. 2017). A subset too small to support the random
# effect returns NULL rather than an unstable fit.
fit <- function(s) {
  if (nrow(s) < 6 || length(unique(s$study)) < 3) return(NULL)
  m <- try(rma.mv(yi, vi, random = ~ 1 | study/obs, data = s, method = "REML"), silent = TRUE)
  if (inherits(m, "try-error")) return(NULL)
  list(est = pct(as.numeric(m$b)), lo = pct(m$ci.lb), hi = pct(m$ci.ub),
       k = m$k, ns = length(unique(s$study)), p = m$pval)
}

# ---- weighted coefficient of variation, for constructing missing dispersion --------------
# Nakagawa et al. (2023): estimate an average coefficient of variation from the observations
# that do report a standard deviation, then use it to construct those that do not. Applied
# identically to both limbs, so the two remain comparable.
cv_pool <- function(m, s, n) {
  ok <- is.finite(m) & is.finite(s) & is.finite(n) & m > 0
  sqrt(sum((n[ok] - 1) * (s[ok] / m[ok])^2) / sum(n[ok] - 1))
}

# ---- log response ratio with the Lajeunesse (2015) second-order correction ---------------
# metafor's ROM measure returns the uncorrected lnRR and its sampling variance; the
# second-order term is added explicitly rather than left implicit, so that the correction is
# visible in the document rather than buried in a package default.
lnRR <- function(mt, st, nt, mc, sc, nc) {
  es <- escalc(measure = "ROM", m1i = mt, sd1i = st, n1i = nt, m2i = mc, sd2i = sc, n2i = nc)
  list(yi = as.numeric(es$yi) + 0.5 * ((st^2)/(nt * mt^2) - (sc^2)/(nc * mc^2)),
       vi = as.numeric(es$vi))
}

# ---- a forest plot, used for Figures 1, 2, 5 and 6 ---------------------------------------
# One drawing routine rather than four, so the four figures cannot drift apart in style.
GREEN <- "#2f6b4f"; RED <- "#9b3b3b"; GREY <- "#6b6b6b"
forest_panel <- function(est, lo, hi, p, labels, notes, xlab,
                         breaks = NULL, lmar = 12.5, group_at = NULL, group_lab = NULL) {
  yy <- seq_along(est)
  xr <- range(c(lo, hi), na.rm = TRUE); xr <- xr + c(-0.06, 0.09) * diff(xr)
  op <- par(mar = c(4.4, lmar, 1.2, 5.2), family = "sans")
  plot(NA, xlim = xr, ylim = c(0.4, length(est) + 0.6), axes = FALSE, xlab = "", ylab = "")
  abline(v = 0, col = "grey55", lwd = 1.2)
  if (!is.null(breaks)) abline(h = breaks, col = "grey85", lwd = 1)
  sig <- p < 0.05
  col <- ifelse(sig, ifelse(est > 0, GREEN, RED), GREY)
  segments(lo, yy, hi, yy, lwd = 2.4, col = col)
  points(est, yy, pch = 21, cex = 1.5, lwd = 1.6, bg = ifelse(sig, col, "white"), col = col)
  axis(1, cex.axis = 0.9); mtext(xlab, side = 1, line = 2.6, cex = 0.95)
  for (i in yy) mtext(labels[i], side = 2, at = i, las = 1, line = 0.4, cex = 0.82, adj = 1)
  for (i in yy) mtext(notes[i], side = 4, at = i, las = 1, line = 0.3, cex = 0.72,
                      adj = 0, col = "grey30")
  if (!is.null(group_at))
    for (j in seq_along(group_at))
      mtext(group_lab[j], side = 2, at = group_at[j], line = lmar - 2.9, cex = 0.88, font = 2)
  par(op)
}

# the compiled published rate estimates, tabulated in Section 2.4 and never pooled
rate <- read.csv("02.inputs/rate-evidence.csv", stringsAsFactors = FALSE)
pa <- rate[rate$study == "Pati 2025", ]
pa_rate  <- pa[pa$quantity == "carbon sequestration rate", ]
pa_stock <- pa[pa$quantity == "total biomass stock", ]
```

# Introduction

Biological invasion is among the recognised direct drivers of biodiversity change, and the international policy response now sets targets for reducing introductions and controlling established invaders [@cbd2022kunming; @roy2024curbing], against costs that have risen steeply [@diagne2021high]. Alongside these concerns, invasion has come to be discussed as a potential contributor to carbon storage, on the reasoning that if invaded systems hold or accumulate more carbon than the communities they replace, invasion presents a trade-off with climate objectives rather than a uniform loss.

That invasion alters ecosystem processes is not in dispute. Invaders change nutrient pools and cycling rates through differences in tissue chemistry, productivity and rooting habit [@ehrenfeld2003effects; @ehrenfeld2010ecosystem], and in some systems reorganise processes such as fire regimes [@dantonio1992biological]. What is disputed is the direction and magnitude of the carbon effect, and the dispute has proved unusually durable.

We argue that its persistence is largely a consequence of conflating two quantities. Carbon *stock* is a mass held at a point in time. Carbon *sequestration rate* is a flux, the mass accumulated per unit area per unit time. The two are related through residence time, and they can move in opposite directions: a system may accumulate carbon rapidly while holding little, if what it accumulates is also lost rapidly. Both are routinely reported as evidence about "carbon sequestration".

The distinction is formalised in carbon accounting. The IPCC defines two approaches to estimating land carbon change and describes them as "two fundamentally different and equally valid approaches": the gain-loss method, which sums annual gains against annual losses, and the stock-difference method, which differences measured stocks at two points in time [@ipcc2006guidelines]. The guidance is explicit that the choice is consequential, noting that the stock-difference method is better suited to large changes and that "where biomass change is very small compared to the total amount of biomass, the inventory error under the stock-difference method may be larger than the expected change" [@ipcc2006guidelines]. Invasion is characteristically that case: a conspicuous annual increment set against a large existing stock.

Read through this lens, the literature is more coherent than it appears. Reports of large positive effects are predominantly about rates and fast-cycling pools. The most-cited synthesis reported pool changes "ranging from a 5% increase in root carbon stock to a 133% increase in shoot C stock", with above-ground production and litter decomposition rising 50 to 120 per cent [@liao2008altered]; every quantified pool in that abstract is a plant pool and its only soil variables are inorganic nitrogen concentrations. Reports of null or negative effects are predominantly about stocks, particularly soil stocks. A synthesis of 1,678 case studies found soil organic carbon greater under native than non-native forests [@lazarolobo2025worldwide]; a synthesis of 2,479 observations found no significant effect of *Spartina alterniflora* on soil organic carbon in vegetated native wetlands [@xu2022effects]; losses have been reported under non-native grasses [@koteen2011invasion; @wilsey2020lower] and in sagebrush steppe [@katz2025invasion].

A mechanism for the divergence is already published. Reviewing invasion effects on forest carbon, @peltzer2010effects concluded that "the direct effects of invaders on forest C are often smaller and shorter-term than their indirect effects caused by altered nutrient availability, primary productivity or species composition, all of which regulate long-term C pools and fluxes", and warned that "without knowing the long-term C trajectories of an invaded system, it will be difficult to demonstrate gains in C as a result of management". A fast-growing invader raises measured rates immediately. Whether that becomes durable stock depends on residence time, governed by slower indirect processes.

Three further considerations bear on measurement. First, soil carbon may be expressed as a concentration or as an areal stock, and the two differ by bulk density; where a vegetation change alters bulk density, a fixed-depth core in denser soil contains more carbon at unchanged concentration [@gifford2003soil; @wendt2013equivalent; @vonhaden2020soils]. The IPCC now states that "with a management induced change in soil bulk density, it is possible to calculate a change in soil carbon stock to a fixed depth in the absence of any change in soil carbon content" [@ipcc2019refinement], although no Tier 1 method corrects for it. Second, soil organic matter is a continuum whose persistence is an ecosystem property [@schmidt2011persistence; @lehmann2015contentious], best resolved into particulate and mineral-associated fractions that differ in residence time [@lavallee2020conceptualizing]; a global radiocarbon synthesis concluded that projections from bulk soil carbon may be misleading [@heckman2022beyond]. Third, vegetation governs the vertical distribution of soil carbon more than the total [@jobbagy2000vertical], and the median depth of recent carbon incorporation is about 10 cm [@balesdent2018atmosphere], so shallow sampling can detect surface gains while missing deeper losses [@fontaine2007stability; @bernal2017invasive].

Finally, invasion effects on carbon are not confined to plants. @peltzer2010effects organise them by trophic level, spanning primary producers, decomposers, herbivores, pathogens, mutualists and predators, and an invasive predator has been shown to alter ecosystem carbon substantially through a trophic cascade with no plant invasion involved [@wardle2007promotion].

The question is not only academic. Where invasion is held to increase carbon storage, the inference drawn is often practical: that invaded land should be retained rather than restored, that invader biomass should be valorised rather than removed, or that fast-growing non-native species are appropriate for carbon plantings. Each of these positions has been argued in print. Invasive *Acacia dealbata* has been proposed as a feedstock whose commercial value could fund its own control [@nunes2021carbon], and *Leucaena* has been argued to build soil carbon in degraded badlands more effectively than native vegetation, through root-derived inputs to mineral-associated pools [@meitha2024invasive]. Against this, @lindenmayer2012avoiding identify the planting of species that become invasive as one of three "bio-perversities" arising from carbon programmes, and @pati2025ecological, having measured rapid carbon accumulation under an invasive tree alongside reduced diversity, title their finding an ecological dilemma. Which of these readings is correct depends on whether the carbon that invasion adds is durable, and durability is a property of stock, not of rate.

We therefore ask four questions. Does invasion increase carbon stocks? Does it increase carbon accumulation rates? Do the two diverge, and does the divergence appear within as well as between studies? And what follows for carbon accounting, which must choose one of two methods that the evidence suggests will disagree.

We made three predictions before analysis. Rate effects would be large and positive; stock effects would be small and centred near zero; and the divergence would be visible within studies reporting both quantities, not only across studies reporting one or the other. We further expected that any stock effect would be concentrated in fast-cycling and aboveground compartments rather than in mineral-associated soil carbon.

# Materials and methods

## 2.1 Two limbs of evidence, treated differently

The two questions require different evidence, and we draw on three bodies of it. Stocks and fluxes are each addressed by a formal meta-analysis of paired invaded and uninvaded observations, fitted by identical methods so that the two limbs are directly comparable. Because no single deposit reports both quantities on the same plots, the two meta-analyses necessarily draw on different primary literatures, and the contrast between them is therefore a between-dataset comparison, with the loss of precision that implies. A third body of evidence, a structured compilation of published carbon accumulation rates, is tabulated and compared but **not** pooled, because too few sources report an accumulation rate against a paired uninvaded control for a pooled estimate to be meaningful. It is used to illustrate the divergence within individual studies, not to estimate its magnitude.

## 2.2 Stock data

Stock effect sizes were extracted from the compilation deposited by @he2025plant (Dryad doi:10.5061/dryad.c59zw3rkc), which assembles paired measurements of soil carbon fractions from invaded plots and adjacent uninvaded controls. This synthesis reuses that deposit rather than conducting a new systematic search; the search, screening and inclusion decisions are those of the source, no new screening was performed, and no PRISMA flow diagram is therefore presented. Reporting conventions otherwise follow @nakagawa2023quantitative and @odea2021preferred. The compilation was chosen because it is, to our knowledge, the only openly deposited invasion dataset carrying multiple soil carbon fractions with paired means, dispersions and sample sizes together with the required moderators.

```{r}
#| label: data-stock
#| echo: true
#| output: false

# ---- read and tidy the stock deposit ----------------------------------------------------
# The workbook carries a two-row header. Columns 1 to 15 hold study metadata; from column 16
# the sheet is a series of blocks, one per carbon fraction, each 12 columns wide, whose name
# sits in row 1 and whose sub-headers sit in row 2:
#   Data source | index | units | Soil depth (cm) | M_native | SD_native | N_native |
#   M_invasive | SD_invasive | N_invasive | lnRR_<fraction> | (blank separator)
raw <- suppressMessages(read_excel("02.inputs/dryad-xu2025/Meta-analysis_Data.xlsx",
                                   col_names = FALSE, .name_repair = "minimal"))
raw <- as.data.frame(raw, stringsAsFactors = FALSE)
r1 <- as.character(unlist(raw[1, ]))    # block names
r2 <- as.character(unlist(raw[2, ]))    # sub-headers
dat <- raw[-(1:2), , drop = FALSE]

META_N <- 15
meta <- dat[, 1:META_N, drop = FALSE]
names(meta) <- c("study", "title", "latitude", "longitude", "methods", "habitat",
                 "native_species", "native_family", "native_growth_cycle",
                 "invasive_species", "invasive_family", "invasive_growth_cycle",
                 "invasion_phase", "MAT", "MAP")

starts <- which(!is.na(r1) & nzchar(r1) & seq_along(r1) > META_N)
frac_code <- sub("^\\s*([A-Za-z0-9]+).*$", "\\1", r1[starts])
SUB <- c("source", "index", "units", "depth", "M_native", "SD_native", "N_native",
         "M_invasive", "SD_invasive", "N_invasive", "lnRR")
num <- function(x) suppressWarnings(as.numeric(ifelse(trimws(as.character(x)) %in%
                                                      c("n/a", "NA", "", "-"), NA, x)))
# the deposit's own lnRR column is deliberately not coerced or used: effect sizes are
# recomputed below, so that both limbs share one estimator and one small-sample correction
VALS <- c("M_native", "SD_native", "N_native", "M_invasive", "SD_invasive", "N_invasive")
out <- list()
for (i in seq_along(starts)) {
  cols <- starts[i] + (0:10)             # 11 populated columns; the 12th is a spacer
  cols <- cols[cols <= ncol(dat)]
  blk <- dat[, cols, drop = FALSE]; names(blk) <- SUB[seq_along(cols)]
  got <- r2[cols]                        # verify the layout rather than assume it
  if (!(grepl("M_native", got[5]) && grepl("SD_native", got[6]) &&
        grepl("M_invasive", got[8]) && grepl("SD_invasive", got[9]))) next
  b <- cbind(meta, fraction = frac_code[i], blk, stringsAsFactors = FALSE)
  for (v in VALS) b[[v]] <- num(b[[v]])
  b$depth_raw <- as.character(blk$depth)
  out[[i]] <- b[is.finite(b$M_native) & is.finite(b$M_invasive), , drop = FALSE]
}
d <- do.call(rbind, out); rownames(d) <- NULL
d$MAT <- num(d$MAT); d$MAP <- num(d$MAP)

# ---- moderators --------------------------------------------------------------------------
# Depth is recorded as a range string. The lower bound drives the depth bands, because the
# prediction under test is that a gain in topsoil masks a loss at depth; the midpoint is the
# continuous form of the same moderator.
dp <- t(vapply(regmatches(d$depth_raw,
                          regexec("^\\s*([0-9.]+)\\s*-\\s*([0-9.]+)", d$depth_raw)),
               function(v) if (length(v) == 3) as.numeric(v[2:3]) else c(NA, NA), c(0, 0)))
d$depth_bot <- dp[, 2]; d$depth_mid <- rowMeans(dp)

h <- tolower(gsub("[^a-z]", "", tolower(d$habitat)))   # labels vary in case and spacing
d$habitat_class <- ifelse(grepl("forest|orchard", h), "forest",
                   ifelse(grepl("grass|hill|plain|plateau|terrace|watershed", h), "grassland",
                   ifelse(grepl("wetland|swamp|marsh|mangrove|delta|estuary|beach|shore|island|river", h),
                          "wetland_coastal",
                   ifelse(grepl("ricefield", h), "cropland", NA))))
d$growth_cycle <- ifelse(grepl("^annual", tolower(d$invasive_growth_cycle)), "annual",
                  ifelse(grepl("perenn", tolower(d$invasive_growth_cycle)), "perennial", NA))
# areal where the reported unit expresses mass per unit ground area, concentration otherwise
AREAL <- c("g/cm2", "kg/m2", "mg ha-1", "mg/ha")
d$basis <- ifelse(d$units %in% AREAL, "areal", "concentration")

# ---- construct missing dispersion, then effect sizes --------------------------------------
cv_nat <- cv_pool(d$M_native, d$SD_native, d$N_native)
cv_inv <- cv_pool(d$M_invasive, d$SD_invasive, d$N_invasive)
d$sd_imputed <- !is.finite(d$SD_native) | !is.finite(d$SD_invasive)
d$SD_native[!is.finite(d$SD_native)]     <- cv_nat * d$M_native[!is.finite(d$SD_native)]
d$SD_invasive[!is.finite(d$SD_invasive)] <- cv_inv * d$M_invasive[!is.finite(d$SD_invasive)]
nmin <- min(c(d$N_native, d$N_invasive), na.rm = TRUE)   # conservative substitute
d$N_native[!is.finite(d$N_native)] <- nmin
d$N_invasive[!is.finite(d$N_invasive)] <- nmin

d <- d[d$M_native > 0 & d$M_invasive > 0, ]
e <- lnRR(d$M_invasive, d$SD_invasive, d$N_invasive, d$M_native, d$SD_native, d$N_native)
d$yi <- e$yi; d$vi <- e$vi
d <- d[is.finite(d$yi) & is.finite(d$vi) & d$vi > 0, ]
d$obs <- seq_len(nrow(d))

soc <- d[d$fraction == "SOC", ]
soc$db <- cut(soc$depth_bot, c(0, 10, 20, 1000), labels = c("0-10 cm", "10-20 cm", ">20 cm"))
n_obs <- nrow(d); n_stud <- length(unique(d$study))
pct_conc <- 100 * mean(soc$basis == "concentration")
n_area_stud <- length(unique(soc$study[soc$basis == "areal"]))
write.csv(d, "03.outputs/he2025-effectsizes.csv", row.names = FALSE)
```

Records were extracted for every fraction reported: total carbon, soil organic carbon, dissolved organic carbon, microbial biomass carbon, particulate organic carbon, mineral-associated organic carbon, readily oxidisable carbon and water-soluble organic carbon. Observations were retained where both means were present and positive, yielding `r f0(n_obs)` paired observations from `r n_stud` studies. Sampling depth was converted to a lower bound and a midpoint; habitats were collapsed into forest, grassland, wetland and coastal, and cropland. The basis of measurement was classified as areal where the reported unit expressed mass per unit ground area and as concentration otherwise, a distinction central to the analysis because the two are related by bulk density.

## 2.3 Flux data

```{r}
#| label: data-flux
#| echo: true
#| output: false

# ---- read and classify the flux deposit ---------------------------------------------------
# The file is not UTF-8. Read as latin1, and the response labels are matched by pattern rather
# than by exact string, because the Greek beta of beta-glucosidase does not survive the
# encoding.
x <- read.csv("02.inputs/dryad-negesse2025/Datasets.csv", stringsAsFactors = FALSE,
              fileEncoding = "latin1", check.names = FALSE)
names(x) <- make.names(names(x))
x$var <- trimws(x$Soil.response.variables)

# Clustering unit. The deposit carries no complete study identifier: the Authors field is blank
# for most rows and Case_studies is a row index, not a source publication. Clustering on either
# would give one group per observation, a degenerate random effect that silently removes the
# correction for non-independence. The site coordinates are present on every row, so the
# coordinate pair is the clustering unit, and the author string is carried forward within each
# site to recover the source publication where one exists.
x$site <- paste(trimws(x$Northing), trimws(x$Easting))
blank <- function(v) is.na(v) | trimws(as.character(v)) %in% c("", "NA", "na")
au <- ifelse(blank(x$Authors), NA_character_, trimws(x$Authors))
x$source <- ave(au, x$site, FUN = function(v) {
  u <- unique(v[!is.na(v)]); if (length(u) == 1) u else NA_character_ })
x$study <- x$site

# ---- functional classes, fixed before any model is fitted ---------------------------------
CARBON_ENZ_PAT <- "glucosidase|Cellulase|Invertase|Glucosamidase|oxidase|Peroxidase|Catalase"
NUTRIENT_ENZ  <- c("Urease", "Acid phosphotase", "Alkaline phosphotase", "Protase",
                   "Glycine amino-peptidase")
NUTRIENT_RATE <- c("N-Mineralization", "N-nitrification")
MICROBIAL_POOL <- c("Microbial biomass", "MBC", "MBN", "Bacteria biomass",
                    "Fungal biomass", "AMF biomass", "Actinomycete biomass")
x$class <- ifelse(x$var == "Soil respiration", "carbon flux",
           ifelse(grepl(CARBON_ENZ_PAT, x$var, ignore.case = TRUE), "carbon-acquiring enzyme",
           ifelse(x$var %in% NUTRIENT_ENZ,   "nutrient-acquiring enzyme",
           ifelse(x$var %in% NUTRIENT_RATE,  "nutrient transformation rate",
           ifelse(x$var %in% MICROBIAL_POOL, "microbial pool", "faunal abundance")))))

nm <- function(v) suppressWarnings(as.numeric(as.character(v)))
x$mc <- nm(x$Mean_Control);   x$sc <- nm(x$SD_Control);   x$nc <- nm(x$N_Control)
x$mt <- nm(x$Mean_Treatment); x$st <- nm(x$SD_Treatment); x$nt <- nm(x$N_Treatment)

# dispersion and effect sizes are constructed exactly as in the stock limb
cv_c <- cv_pool(x$mc, x$sc, x$nc); cv_t <- cv_pool(x$mt, x$st, x$nt)
x$sd_imputed <- !is.finite(x$sc) | !is.finite(x$st)
x$sc[!is.finite(x$sc)] <- cv_c * x$mc[!is.finite(x$sc)]
x$st[!is.finite(x$st)] <- cv_t * x$mt[!is.finite(x$st)]
nmin_f <- min(c(x$nc, x$nt), na.rm = TRUE)
x$nc[!is.finite(x$nc)] <- nmin_f; x$nt[!is.finite(x$nt)] <- nmin_f

x <- x[is.finite(x$mc) & is.finite(x$mt) & x$mc > 0 & x$mt > 0, ]
ef <- lnRR(x$mt, x$st, x$nt, x$mc, x$sc, x$nc)
x$yi <- ef$yi; x$vi <- ef$vi
x <- x[is.finite(x$yi) & is.finite(x$vi) & x$vi > 0, ]
x$obs <- seq_len(nrow(x))

n_flux <- nrow(x); n_site <- length(unique(x$study))
n_ctry <- length(unique(x$Country)); n_src <- length(unique(na.omit(x$source)))
pct_imp_flux <- 100 * mean(x$sd_imputed)
```

Flux effect sizes were extracted from the compilation deposited by @negesse2025invasive (Dryad doi:10.5061/dryad.hhmgqnkq5), which assembles paired measurements of soil processes from invaded plots and adjacent uninvaded controls. As with the stock limb, the search, screening and inclusion decisions are those of the source. Retaining observations with positive paired means yielded `r f0(n_flux)` paired observations from `r n_site` sites in `r n_ctry` countries on six continents.

The deposit carries no complete study identifier: the author field is populated for only a minority of records and its case-study field is a row index rather than a source publication. Site coordinates, by contrast, are recorded for every observation and resolve to `r n_site` distinct locations, so the coordinate pair was used as the clustering unit in all models and the author string was carried forward within each site, recovering a named source publication for `r n_src` of the `r n_site` sites. Clustering on site rather than publication is conservative where one publication reports several sites, since it treats those sites as independent, and we note this as a limitation of the deposit rather than a choice.

Response variables were assigned to six functional classes before any model was fitted: soil respiration, which is a direct carbon flux and the quantity of primary interest; carbon-acquiring extracellular enzymes; nutrient-acquiring extracellular enzymes; nitrogen transformation rates; microbial pools; and faunal abundance. Carbon-acquiring and nutrient-acquiring enzymes were separated because the two groups index different limitations, the first the supply of degradable carbon and the second the demand for nitrogen and phosphorus, and an invader that accelerates carbon cycling specifically should raise the first more than the second. Keeping them apart makes that a testable prediction rather than an assumption. Enzyme activities and microbial pools are not fluxes, and we do not treat them as such; they are the proximate machinery of decomposition and are reported as evidence on the mechanism.

Effect sizes were recomputed from the reported means, dispersions and sample sizes rather than taken from the deposit's own log ratio column, so that both limbs use identical estimators and identical corrections. Every flux observation reported a standard deviation and a sample size, so dispersion was constructed for `r f1(pct_imp_flux)`% of them.

## 2.4 Rate data

Rate estimates were compiled from studies reporting a carbon accumulation rate, net primary production, or a proportional change in a carbon pool attributable to invasion, alongside an uninvaded comparison. Entries were admitted only where the uninvaded comparison was explicit, which excludes the large literature reporting invaded productivity without a control. Each entry records the invader, its trophic level following the classification of @peltzer2010effects, the quantity measured, the invaded and uninvaded values with dispersion where the source reports it, the ratio, and the source.

The compilation is deliberately conservative in four respects. Entries are reproduced exactly as published, without recalculation. A reported range is recorded as a range and plotted as an interval rather than reduced to a midpoint. No dispersion is imputed, so entries lacking it are plotted as points without intervals and are not weighted. And where a single study reports both a rate and a stock, both are entered, because such studies are the strongest available test of the divergence and their internal comparison is free of between-study confounding.

One source considered for inclusion was excluded on quality grounds. A review reporting that *Phragmites*-invaded wetlands show 3.1 times higher net primary production is published in a secondary-school student journal, carries no digital object identifier, and attributes the figure to a further source rather than to original measurement. It is not cited here, and the underlying claim should be traced to its primary source before use.

Invader trophic level was assigned following @peltzer2010effects, who distinguish primary producers, decomposers, herbivores, plant pathogens, mutualists and predators, and who argue that each group influences ecosystem carbon through a characteristic combination of direct and indirect pathways. The classification matters because it tests whether the rate-stock divergence is a property of fast-growing plants specifically or of invasion more generally.

```{r}
#| label: tbl-rate
#| echo: true
#| tbl-cap: "Published estimates of the effect of invasion on carbon accumulation rates and on the corresponding stocks. Values are reproduced as published. Where dispersion is reported it is a standard error. This compilation is tabulated rather than pooled: four sources cannot support a pooled estimate, and the entries differ in quantity and in units."

t_rate <- data.frame(
  Source = rate$study, Invader = rate$invader, `Trophic level` = rate$trophic_level,
  Quantity = rate$quantity,
  Invaded = ifelse(is.na(rate$invaded_err) | rate$invaded_err == "",
                   as.character(rate$invaded),
                   sprintf("%s ± %s", rate$invaded, rate$invaded_err)),
  Uninvaded = ifelse(is.na(rate$uninvaded) | rate$uninvaded == "", "-",
                     sprintf("%s ± %s", rate$uninvaded, rate$uninvaded_err)),
  `Ratio` = sprintf("%.2f", rate$ratio),
  check.names = FALSE, stringsAsFactors = FALSE)
kable(t_rate, row.names = FALSE)
write.csv(t_rate, file.path(repo, "03.outputs/tables/table-rate-evidence.csv"), row.names = FALSE)
```

## 2.5 Effect sizes and models

Stock effects were quantified as the log response ratio, the natural logarithm of the ratio of invaded to uninvaded means [@hedges1999meta], which is scale-invariant and therefore permits observations reported in ten different units to be combined. The second-order small-sample bias correction of @lajeunesse2015bias was applied explicitly. Where a standard deviation was absent it was constructed from a weighted average coefficient of variation estimated across observations reporting dispersion, following @nakagawa2023robust; where sample size was absent the smallest observed value was substituted, which inflates the sampling variance and downweights the observation.

Pooled effects were estimated with multilevel random-effects models fitted by restricted maximum likelihood in `metafor` [@viechtbauer2010conducting], with observations nested within studies to account for non-independence [@noble2017nonindependence]. Heterogeneity was quantified as $I^2$ following @nakagawa2012methodological. Moderators tested on the soil organic carbon subset were sampling depth, habitat class, invader growth cycle, temperature and precipitation, with the basis of measurement as the moderator of primary interest. Small-study effects were assessed by regressing effect size on its standard error. Robustness was examined by refitting after excluding imputed dispersion, with equal weighting, and as a single-level model.

## 2.6 Turnover classification

The interpretation that invasion accelerates cycling rather than building durable stock makes a testable prediction: the effect of invasion on a carbon pool should be largest in the fastest-cycling pools and should decline toward zero in the slowest. To test it, each measured fraction was assigned to a turnover class from its published characterisation, before any model was fitted and without reference to the results.

Microbial biomass, dissolved organic, water-soluble organic and readily oxidisable carbon were classified as **fast**, turning over on timescales of days to months and comprising living or freely soluble material. Particulate organic carbon, composed of partially decomposed plant residues, was classified as **intermediate**, on timescales of years to decades. Mineral-associated organic carbon was classified as **slow**, on timescales of decades to centuries, following the characterisation of these two fractions as fundamentally different in formation and persistence [@lavallee2020conceptualizing] and the radiocarbon evidence that they differ consistently in persistence [@heckman2022beyond]. Bulk soil organic carbon and total carbon aggregate all of these and were classified as mixed and excluded from the ordered test.

The prediction was tested by entering turnover rank as a continuous moderator, which tests the ordered hypothesis directly rather than testing for any difference among classes, and confirmed with a categorical specification.

# Results

## 3.1 Carbon stocks show no reliable effect

```{r}
#| label: fit-fractions
#| echo: true

# Pooled effect of invasion on each carbon fraction. I-squared for a multilevel model follows
# Nakagawa & Santos (2012): the ratio of total heterogeneity variance to that plus the typical
# within-study sampling variance.
FRACS <- c("SOC", "TC", "POC", "MAOC", "MBC", "DOC", "ROC", "WSOC")
fit_frac <- function(k) {
  sset <- d[d$fraction == k, ]
  if (length(unique(sset$study)) < 3) return(NULL)
  m <- try(rma.mv(yi, vi, random = ~ 1 | study/obs, data = sset, method = "REML"), silent = TRUE)
  if (inherits(m, "try-error")) return(NULL)
  W <- diag(1 / sset$vi); X <- model.matrix(m)
  P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
  data.frame(fraction = k, k_obs = m$k, k_study = length(unique(sset$study)),
             pct = pct(as.numeric(m$b)), pct_lb = pct(m$ci.lb), pct_ub = pct(m$ci.ub),
             pval = m$pval,
             I2 = 100 * sum(m$sigma2) / (sum(m$sigma2) + (m$k - m$p) / sum(diag(P))),
             stringsAsFactors = FALSE)
}
res <- do.call(rbind, lapply(FRACS, fit_frac))
gf <- function(k) as.list(res[res$fraction == k, ])
r_mbc <- gf("MBC"); r_poc <- gf("POC"); r_maoc <- gf("MAOC")
r_soc  <- fit(soc)
r_conc <- fit(soc[soc$basis == "concentration", ]); r_area <- fit(soc[soc$basis == "areal", ])
m_unw  <- rma.mv(yi, rep(1, nrow(soc)), random = ~ 1 | study/obs, data = soc, method = "REML")
write.csv(res, file.path(TBL, "pooled-by-fraction.csv"), row.names = FALSE)

kable(data.frame(Fraction = res$fraction, Observations = res$k_obs, Studies = res$k_study,
                 `Change (%)` = sg(res$pct),
                 `95% CI` = sprintf("%s to %s", sg(res$pct_lb), sg(res$pct_ub)),
                 p = signif(res$pval, 3), `I2 (%)` = round(res$I2),
                 check.names = FALSE), row.names = FALSE)
```

```{r}
#| label: fig-fractions
#| echo: true
#| fig-width: 7.3
#| fig-height: 5.0
#| fig-cap: "Pooled effect of invasion on eight soil carbon fractions. Points are percentage changes from multilevel random-effects models with 95% confidence intervals; filled symbols indicate intervals excluding zero. Numbers at right give observations and, in parentheses, contributing studies."

LB <- c(MBC = "Microbial biomass C", SOC = "Soil organic C", MAOC = "Mineral-associated C",
        DOC = "Dissolved organic C", ROC = "Readily oxidisable C", TC = "Total C",
        WSOC = "Water-soluble organic C", POC = "Particulate organic C")
o1 <- res[order(res$pct), ]
draw_fig1 <- function()
  forest_panel(o1$pct, o1$pct_lb, o1$pct_ub, o1$pval, LB[o1$fraction],
               sprintf("k=%d (%d)", o1$k_obs, o1$k_study),
               "Change under invasion (%)")
png(file.path(FIG, "figure-1-fractions.png"), width = 2200, height = 1500, res = 300)
draw_fig1(); invisible(dev.off())    # archived copy, reproduced in the repository README
draw_fig1()
```

Soil organic carbon was `r sg(r_soc$est)`% higher under invasion, but the interval spanned zero (95% CI `r sg(r_soc$lo)` to `r sg(r_soc$hi)`%, *p* `r pf(r_soc$p)`), and the estimate was not robust to specification, falling to `r sg(pct(as.numeric(m_unw$b)))`% under equal weighting. Of the eight fractions examined, only two had intervals excluding zero: microbial biomass carbon rose `r sg(r_mbc$pct)`% (*p* `r pf(r_mbc$pval)`) and particulate organic carbon fell `r f1(abs(r_poc$pct))`% (*p* `r pf(r_poc$pval)`). Mineral-associated organic carbon, the fraction most relevant to durable storage, rested on `r r_maoc$k_obs` observations from `r r_maoc$k_study` studies and could not be resolved. Heterogeneity was at least `r f1(min(res$I2))`% in every fraction.

```{r}
#| label: fit-moderators
#| echo: true

# Moderator tests on the soil organic carbon subset. Each is a separate model rather than one
# multi-moderator model, because the moderators are unevenly observed and a single model would
# silently drop every row missing any one of them.
mod_test <- function(form, label) {
  dd <- soc[complete.cases(soc[, all.vars(form), drop = FALSE]), ]
  if (nrow(dd) < 10 || length(unique(dd$study)) < 3) return(NULL)
  m <- try(rma.mv(yi, vi, mods = form, random = ~ 1 | study/obs, data = dd, method = "REML"),
           silent = TRUE)
  if (inherits(m, "try-error")) return(NULL)
  data.frame(Moderator = label, QM = m$QM, df = m$m, p = m$QMp, n = nrow(dd),
             stringsAsFactors = FALSE)
}
mods <- do.call(rbind, list(
  mod_test(~ depth_mid,             "Soil depth (midpoint)"),
  mod_test(~ factor(habitat_class), "Habitat class"),
  mod_test(~ factor(growth_cycle),  "Invader growth cycle"),
  mod_test(~ MAT,                   "Mean annual temperature"),
  mod_test(~ MAP,                   "Mean annual precipitation"),
  mod_test(~ factor(basis),         "Basis of measurement")))
write.csv(mods, file.path(TBL, "soc-moderators.csv"), row.names = FALSE)

m_hab  <- rma.mv(yi, vi, mods = ~ factor(habitat_class), random = ~ 1 | study/obs,
                 data = soc[!is.na(soc$habitat_class), ], method = "REML")
r_wet  <- fit(soc[which(soc$habitat_class == "wetland_coastal"), ])
r_for  <- fit(soc[which(soc$habitat_class == "forest"), ])
r_gra  <- fit(soc[which(soc$habitat_class == "grassland"), ])
r_shal <- fit(soc[which(soc$db == "0-10 cm"), ])
r_mid  <- fit(soc[which(soc$db == "10-20 cm"), ])
r_deep <- fit(soc[which(soc$db == ">20 cm"), ])

kable(data.frame(Moderator = mods$Moderator, `Q_M` = round(mods$QM, 2), df = mods$df,
                 p = signif(mods$p, 3), Observations = mods$n, check.names = FALSE),
      row.names = FALSE)
```

```{r}
#| label: fig-moderators
#| echo: true
#| fig-width: 7.3
#| fig-height: 5.6
#| fig-cap: "Effect of invasion on soil organic carbon by habitat class, sampling depth and basis of measurement, from multilevel models fitted separately to each subset."

G <- list(list("Areal stock", r_area, 3), list("Concentration", r_conc, 3),
          list(">20 cm", r_deep, 2), list("10-20 cm", r_mid, 2), list("0-10 cm", r_shal, 2),
          list("Forest", r_for, 1), list("Grassland", r_gra, 1),
          list("Wetland and coastal", r_wet, 1))
G <- G[!vapply(G, function(g) is.null(g[[2]]), TRUE)]
gl <- vapply(G, function(g) g[[1]], ""); gp <- vapply(G, function(g) g[[3]], 0)
ge <- vapply(G, function(g) g[[2]]$est, 0); glo <- vapply(G, function(g) g[[2]]$lo, 0)
ghi <- vapply(G, function(g) g[[2]]$hi, 0); gpv <- vapply(G, function(g) g[[2]]$p, 0)
gk  <- vapply(G, function(g) g[[2]]$k, 0);  gns <- vapply(G, function(g) g[[2]]$ns, 0)
draw_fig2 <- function()
  forest_panel(ge, glo, ghi, gpv, gl, sprintf("k=%d (%d)", gk, gns),
               "Change in soil organic carbon under invasion (%)",
               breaks = which(diff(gp) != 0) + 0.5, lmar = 11.5,
               group_at = tapply(seq_along(gp), gp, mean),
               group_lab = c(`1` = "Habitat", `2` = "Depth", `3` = "Measurement")[
                 names(tapply(seq_along(gp), gp, mean))])
png(file.path(FIG, "figure-2-moderators.png"), width = 2200, height = 1700, res = 300)
draw_fig2(); invisible(dev.off())
draw_fig2()
```

Habitat was a significant moderator (*Q*~M~ = `r f1(m_hab$QM)`, df = `r m_hab$m`, *p* `r pf(m_hab$QMp)`). The response was large and positive in wetland and coastal systems (`r ci(r_wet)`) and indistinguishable from zero in forest (`r ci(r_for)`) and grassland (`r ci(r_gra)`). The response declined with depth, from `r ci(r_shal)` in the uppermost 10 cm to `r ci(r_deep)` below 20 cm.

Decisively, `r f1(pct_conc)`% of soil organic carbon observations were reported as concentrations rather than areal stocks. Among concentration observations the pooled effect was `r ci(r_conc)`; among the `r r_area$k` areal observations, from `r n_area_stud` studies, it was `r ci(r_area)`. No contributing study reported bulk density alongside concentration, so the concentration observations cannot be converted to stocks.

```{r}
#| label: coverage
#| echo: true

# The pooled estimates are only as general as the studies behind them, so the composition of
# the evidence base is reported as a result rather than left implicit. Species names in the
# source run words together, carry trailing authorities and contain transcription errors, so
# they are mapped explicitly rather than reformatted by rule: every correction stays visible.
SPECIES_MAP <- c(
  "S.alterniflora" = "Spartina alterniflora", "Phyllostachysedulis" = "Phyllostachys edulis",
  "P.australis" = "Phragmites australis",
  "Leucaeleucocephala(Lam.)deWit" = "Leucaena leucocephala",
  "Prosopisjuliflora (Sw.)DC." = "Prosopis juliflora", "Bromustectorum L." = "Bromus tectorum",
  "Mikaniamicrantha KunthinHumb.&al." = "Mikania micrantha",
  "Ipomoeacairica (L.)Sweet)" = "Ipomoea cairica",
  "Sphagneticolatrilobata(L.)Pruski" = "Sphagneticola trilobata",
  "Chromolaeodorata (L.)R.M.King&H.Rob.)" = "Chromolaena odorata",
  "Chromolaeodorata L." = "Chromolaena odorata",
  "AmbrosiaartemisiifoliaL." = "Ambrosia artemisiifolia",
  "Sonneratiaapetala Buch.-Ham." = "Sonneratia apetala",
  "Robiniapseudoacacia L." = "Robinia pseudoacacia", "Bidenspilosa L." = "Bidens pilosa",
  "Polygonumcuspidatum" = "Reynoutria japonica", "Puerariamonta" = "Pueraria montana",
  "Praxelisclematidea" = "Praxelis clematidea", "A.fruticosa" = "Amorpha fruticosa")
# Chromolaena is misspelled "Chromolaeo" throughout the source; Polygonum cuspidatum is the
# basionym of Reynoutria japonica and is standardised to the accepted name. Matching uses a
# normalised key, letters only and lower case, because the source varies in spacing and
# authority citation for the same taxon.
norm_key <- function(v) tolower(gsub("[^A-Za-z]", "", v))
KEY_MAP <- setNames(SPECIES_MAP, norm_key(names(SPECIES_MAP)))
tidy_sp <- function(v) {
  hit <- unname(KEY_MAP[norm_key(v)])
  fallback <- trimws(gsub("\\s*\\(.*?\\)|[A-Z]\\.[A-Za-z]*&.*$|\\s+[A-Z][a-z]*\\.?$", "", v))
  ifelse(is.na(hit), fallback, hit)
}
soc$species <- tidy_sp(soc$invasive_species)
soc$family  <- trimws(gsub("[A-Z][a-z]*\\.\\s*&.*$|Bercht.*$|Juss\\.$", "", soc$invasive_family))

sp <- sort(table(soc$species), decreasing = TRUE)
top <- data.frame(Species = names(sp), Observations = as.integer(sp),
                  `Share (%)` = round(100 * as.integer(sp) / nrow(soc), 1),
                  Studies = vapply(names(sp), function(k)
                    length(unique(soc$study[soc$species == k])), 0L),
                  check.names = FALSE)
fm <- sort(table(soc$family), decreasing = TRUE)
famtab <- data.frame(Family = names(fm), Observations = as.integer(fm),
                     `Share (%)` = round(100 * as.integer(fm) / nrow(soc), 1),
                     check.names = FALSE)
write.csv(top, file.path(TBL, "table-coverage-species.csv"), row.names = FALSE)
write.csv(famtab, file.path(TBL, "table-coverage-family.csv"), row.names = FALSE)
top_share <- top$`Share (%)`[1]
poaceae_share <- famtab$`Share (%)`[famtab$Family == "Poaceae"][1]

kable(head(top, 8), row.names = FALSE)
```

The evidence base is also narrow. Of the soil organic carbon observations, `r f1(top_share)`% concern a single species, *Spartina alterniflora*, and `r f1(poaceae_share)`% concern the grass family alone.

## 3.2 Carbon fluxes and the processes driving them rise

```{r}
#| label: fit-flux
#| echo: true

# Pooled effect of invasion within each functional class, and the two contrasts the argument
# makes testable: whether carbon-acquiring enzymes respond differently from nutrient-acquiring
# enzymes, and whether the result holds within forests alone.
CLS <- c("carbon flux", "nutrient transformation rate", "nutrient-acquiring enzyme",
         "carbon-acquiring enzyme", "microbial pool", "faunal abundance")
fit_class <- function(k) {
  sset <- x[x$class == k, ]
  f <- fit(sset); if (is.null(f)) return(NULL)
  data.frame(class = k, k_obs = f$k, k_study = f$ns, pct = f$est,
             lo = f$lo, hi = f$hi, p = f$p, stringsAsFactors = FALSE)
}
fcls <- do.call(rbind, lapply(CLS, fit_class))
gc_ <- function(k) as.list(fcls[fcls$class == k, ])
q_resp <- gc_("carbon flux");               q_cenz  <- gc_("carbon-acquiring enzyme")
q_nenz <- gc_("nutrient-acquiring enzyme"); q_pool  <- gc_("microbial pool")
q_faun <- gc_("faunal abundance");          q_nrate <- gc_("nutrient transformation rate")
write.csv(fcls, file.path(TBL, "table-flux-classes.csv"), row.names = FALSE)

sr <- x[x$class == "carbon flux", ]
m_enz <- rma.mv(yi, vi, mods = ~ factor(class), random = ~ 1 | study/obs,
                data = x[x$class %in% c("carbon-acquiring enzyme",
                                        "nutrient-acquiring enzyme"), ], method = "REML")
fo  <- x[trimws(x$Forest.vs.Non.forest) == "Forest", ]
frs <- fo[fo$class == "carbon flux", ]
m_frs <- rma.mv(yi, vi, random = ~ 1 | study/obs, data = frs, method = "REML")
n_forest_resp_site <- length(unique(frs$study))

kable(data.frame(Class = fcls$class, Observations = fcls$k_obs, Sites = fcls$k_study,
                 `Change (%)` = sg(fcls$pct),
                 `95% CI` = sprintf("%s to %s", sg(fcls$lo), sg(fcls$hi)),
                 p = signif(fcls$p, 3), check.names = FALSE), row.names = FALSE)
```

The flux limb reverses the stock result. Across `r f0(n_flux)` paired observations from `r n_site` sites, soil respiration was `r cif(q_resp)` higher under invasion (*p* `r pf(q_resp$p)`), on `r q_resp$k_obs` observations from `r q_resp$k_study` sites.



```{r}
#| label: fig-flux
#| echo: true
#| fig-width: 7.6
#| fig-height: 4.8
#| fig-cap: "Pooled effect of invasion on soil processes, by functional class. Points are percentage changes from multilevel random-effects models with 95% confidence intervals, with observations clustered by site; filled symbols indicate intervals excluding zero. Numbers at right give observations and, in parentheses, contributing sites."

o6 <- fcls[order(fcls$pct), ]
draw_fig6 <- function()
  forest_panel(o6$pct, o6$lo, o6$hi, o6$p, o6$class,
               sprintf("k=%d (%d)", o6$k_obs, o6$k_study),
               "Change under invasion (%)", lmar = 13.5)
png(file.path(FIG, "figure-6-flux.png"), width = 2300, height = 1450, res = 300)
draw_fig6(); invisible(dev.off())
draw_fig6()
```

The machinery of decomposition moved with it. Carbon-acquiring extracellular enzyme activity rose `r cif(q_cenz)` (*p* `r pf(q_cenz$p)`) and microbial biomass `r cif(q_pool)` (*p* `r pf(q_pool$p)`). Soil fauna moved the other way, falling `r f1(abs(q_faun$pct))`% (*p* `r pf(q_faun$p)`). Nitrogen transformation rates were not resolved (`r cif(q_nrate)`).

Two results qualify this. Nutrient-acquiring enzymes rose `r cif(q_nenz)`, statistically indistinguishable from the carbon-acquiring group (*Q*~M~ = `r f1(m_enz$QM)`, df = `r m_enz$m`, *p* `r pf(m_enz$QMp)`), so the acceleration is not specific to carbon acquisition and is better read as a general intensification of microbial activity. And on the forest subset alone, soil respiration was `r sg(pct(as.numeric(m_frs$b)))`% (*p* `r pf(m_frs$pval)`) on `r m_frs$k` observations from `r n_forest_resp_site` sites, which is far too few to resolve; the flux result is a cross-biome one and is not demonstrated within forests specifically.

## 3.3 Published accumulation rates point the same way

Published rate estimates diverge from the stock results in the same direction (Table 1). In a tropical dry deciduous secondary forest of Central India, invasion by *Gliricidia sepium* was associated with a carbon sequestration rate of `r pa_rate$invaded` ± `r pa_rate$invaded_err` Mg C ha^-1^ yr^-1^ against `r pa_rate$uninvaded` ± `r pa_rate$uninvaded_err` in uninvaded forest, a `r f1(pa_rate$ratio)`-fold difference [@pati2025ecological]. A global synthesis reported fluxes including above-ground production and litter decomposition rising 50 to 120 per cent, and shoot carbon rising 133 per cent [@liao2008altered].

```{r}
#| label: fig-ratestock
#| echo: true
#| fig-width: 8.0
#| fig-height: 5.5
#| fig-cap: "The rate-stock divergence. Estimates of the effect of invasion on carbon, expressed as the ratio of invaded to uninvaded values on a logarithmic axis. Diamonds are published rate or flux estimates, which are shown individually and are not pooled. Circles are stock estimates, either fitted here from the meta-analysis with 95% confidence intervals or reproduced from published sources. Sources are given at right."

# Published rate entries are individual estimates and are deliberately NOT pooled; the panel
# says so, so that the figure cannot be read as a pooled comparison. A reported range is one
# estimate plotted as an interval, not two points.
R <- data.frame(label = character(), lr = numeric(), lo = numeric(), hi = numeric(),
                cls = character(), src = character(), stringsAsFactors = FALSE)
add <- function(label, lr, lo = NA, hi = NA, cls, src)
  R <<- rbind(R, data.frame(label, lr, lo, hi, cls, src, stringsAsFactors = FALSE))
LABS <- c("carbon sequestration rate" = "Sequestration rate, Gliricidia",
          "live plant biomass carbon" = "Live plant biomass C, rats",
          "non-living pool carbon"    = "Non-living pool C, rats",
          "total ecosystem carbon"    = "Total ecosystem C, rats",
          "shoot carbon stock"        = "Shoot C, global synthesis",
          "root carbon stock"         = "Root C, global synthesis")
for (i in seq_len(nrow(rate))) {
  q <- rate$quantity[i]
  if (q == "total biomass stock" || rate$err_type[i] == "range") next
  lab <- if (q %in% names(LABS)) LABS[[q]] else q
  add(lab, log(rate$ratio[i]), NA, NA,
      if (grepl("Sequestration rate|ANPP", lab)) "rate or flux" else "stock", rate$study[i])
}
rng <- rate[rate$err_type == "range", ]
if (nrow(rng) == 2)
  add("ANPP and decomposition, global", mean(log(rng$ratio)), min(log(rng$ratio)),
      max(log(rng$ratio)), "rate or flux", rng$study[1])

lg <- function(f) c(log(1 + f$est/100), log(1 + f$lo/100), log(1 + f$hi/100))
for (nm in list(c("Soil organic C, all observations", "all"),
                c("Soil organic C, concentration", "conc"),
                c("Soil organic C, areal stock", "areal"))) {
  f <- switch(nm[2], all = r_soc, conc = r_conc, areal = r_area)
  v <- lg(f); add(nm[1], v[1], v[2], v[3], "stock", "this study")
}
pb <- rate[rate$quantity == "total biomass stock", ]
add("Biomass stock, Gliricidia", log(pb$ratio), NA, NA, "stock", pb$study)
for (k in c("MBC", "POC")) {
  r <- res[res$fraction == k, ]
  add(paste0(ifelse(k == "MBC", "Microbial biomass C", "Particulate organic C"), ", fractions"),
      log(1 + r$pct/100), log(1 + r$pct_lb/100), log(1 + r$pct_ub/100), "stock", "this study")
}
# soil respiration, the flux limb's headline, plotted on the same axis
add("Soil respiration, flux limb", log(1 + q_resp$pct/100), log(1 + q_resp$lo/100),
    log(1 + q_resp$hi/100), "rate or flux", "this study")
R <- R[order(R$cls, R$lr), ]

draw_fig4 <- function() {
  op <- par(mar = c(4.6, 15.5, 2.2, 5.6), family = "sans")
  yy <- seq_len(nrow(R))
  xr <- range(c(R$lr, R$lo, R$hi), na.rm = TRUE); xr <- xr + c(-0.12, 0.12) * diff(xr)
  plot(NA, xlim = xr, ylim = c(0.3, nrow(R) + 0.7), axes = FALSE, xlab = "", ylab = "")
  abline(v = 0, col = "grey50", lwd = 1.3)
  abline(h = which(diff(as.numeric(factor(R$cls))) != 0) + 0.5, col = "grey85", lwd = 1)
  isr <- R$cls == "rate or flux"
  segments(R$lo, yy, R$hi, yy, lwd = 2.2, col = ifelse(isr, GREEN, GREY))
  points(R$lr, yy, pch = ifelse(isr, 23, 21), cex = 1.5, lwd = 1.5,
         bg = ifelse(isr, GREEN, "white"), col = ifelse(isr, GREEN, GREY))
  tk <- log(c(0.5, 1, 2, 4, 8, 16)); tk <- tk[tk >= xr[1] & tk <= xr[2]]
  axis(1, at = tk, labels = sprintf("%g", exp(tk)), cex.axis = 0.9)
  mtext("Invaded / uninvaded ratio (log scale)", side = 1, line = 2.7, cex = 0.95)
  for (i in yy) mtext(R$label[i], side = 2, at = i, las = 1, line = 0.4, cex = 0.78, adj = 1)
  for (i in yy) mtext(R$src[i], side = 4, at = i, las = 1, line = 0.3, cex = 0.68,
                      adj = 0, col = "grey35")
  pl <- tapply(yy, R$cls, mean)
  mtext("Rate or flux", side = 2, at = pl[["rate or flux"]], line = 13.2, cex = 0.85, font = 2)
  mtext("Stock", side = 2, at = pl[["stock"]], line = 13.2, cex = 0.85, font = 2)
  legend("bottomright", bty = "n", cex = 0.72, inset = c(0, 0.02),
         pch = c(23, 21), pt.bg = c(GREEN, "white"), col = c(GREEN, GREY),
         legend = c("rate or flux (published, not pooled)", "stock (fitted or published)"))
  par(op)
}
png(file.path(FIG, "figure-4-rate-stock.png"), width = 2400, height = 1650, res = 300)
draw_fig4(); invisible(dev.off())
draw_fig4()
```

Plotted on a common ratio scale (Figure 4), the two classes of estimate separate. Rate and flux estimates range from `r f1(exp(0.4055))`-fold to `r f1(pa_rate$ratio)`-fold. Stock estimates, whether fitted here or drawn from published sources, fall between `r f1(exp(-0.4521))`-fold and `r f1(exp(0.8459))`-fold, and the two stock estimates with the largest magnitude lie in opposite directions: particulate organic carbon below unity and shoot carbon above it.

The stock class is itself structured. Aboveground plant pools rise: shoot carbon by 133% in a global synthesis [@liao2008altered] and live plant biomass carbon by 104% under invasive predators [@wardle2007promotion]. Soil pools do not, and the non-living pools in the same island system fall by 26%. The distinction between plant and soil stocks therefore runs alongside the distinction between rate and stock, and both point the same way: the carbon that invasion adds is held in fast-cycling, aboveground or labile compartments.

## 3.4 The divergence tested directly, within and between datasets

The clearest demonstration is internal to one study. @pati2025ecological measured both quantities on the same plots: total biomass was `r pa_stock$invaded` ± `r pa_stock$invaded_err` Mg ha^-1^ under invasion against `r pa_stock$uninvaded` ± `r pa_stock$uninvaded_err` uninvaded, a `r f1(pa_stock$ratio)`-fold difference, while the sequestration rate differed `r f1(pa_rate$ratio)`-fold. A stock comparison and a rate comparison on the same plots in the same year therefore differ by nearly an order of magnitude in the apparent size of the invasion effect. The uninvaded rate additionally has a standard error exceeding its mean and is not distinguishable from zero, so the ratio should be read as a lower bound on the contrast rather than a precise multiple.

```{r}
#| label: fit-contrast
#| echo: true

# Soil respiration and soil carbon stock are fitted in one model with a limb indicator as
# moderator. The stock limb is restricted to areal observations, because a concentration is not
# a stock: carbon per unit mass of soil leaves the bulk density change under invasion
# uncontrolled, and that is precisely the confound the areal restriction removes. The other two
# specifications are reported alongside it so the effect of the restriction is visible.
contrast <- function(sub) {
  st <- data.frame(yi = sub$yi, vi = sub$vi, study = paste0("S_", sub$study),
                   limb = "stock", stringsAsFactors = FALSE)
  fl <- data.frame(yi = sr$yi, vi = sr$vi, study = paste0("F_", sr$study),
                   limb = "flux", stringsAsFactors = FALSE)
  both <- rbind(st, fl); both$obs <- seq_len(nrow(both))
  rma.mv(yi, vi, mods = ~ factor(limb), random = ~ 1 | study/obs, data = both, method = "REML")
}
ca      <- contrast(soc[soc$basis == "areal", ])
c_conc  <- contrast(soc[soc$basis == "concentration", ])
c_all   <- contrast(soc)
ca_gap  <- pct(-ca$b[2])        # by how much the flux effect exceeds the areal stock effect

t_con <- data.frame(
  `Stock limb` = c("Areal stock (primary)", "Concentration", "All stock observations"),
  `Gap (pp)` = sg(c(ca_gap, pct(-c_conc$b[2]), pct(-c_all$b[2]))),
  `Q_M` = round(c(ca$QM, c_conc$QM, c_all$QM), 2),
  p = signif(c(ca$QMp, c_conc$QMp, c_all$QMp), 3),
  `Stock obs` = c(r_area$k, r_conc$k, r_soc$k), check.names = FALSE)
write.csv(t_con, file.path(TBL, "table-flux-stock-contrast.csv"), row.names = FALSE)
kable(t_con, row.names = FALSE)
```

Tested formally across datasets, the divergence is in the predicted direction but does not reach significance. Fitting soil respiration and areal soil carbon in one model with a limb indicator, the flux response exceeded the stock response by `r sg(ca_gap)` percentage points (*Q*~M~ = `r f1(ca$QM)`, df = `r ca$m`, *p* `r pf(ca$QMp)`). The gap is of the size the argument predicts and is the largest of the three specifications we fitted, but it rests on `r ca$k - r_area$k` flux observations against `r r_area$k` areal stock observations, and with that many observations an effect of this size would not be expected to separate from zero. Against concentration observations, which are not stocks, the gap narrows to `r sg(pct(-c_conc$b[2]))` percentage points (*p* `r pf(c_conc$QMp)`): the flux response and the concentration response are nearly identical, which is what should happen if a rise in carbon concentration under invasion reflects the same intensification of cycling that raises respiration, rather than accrual of stock.

We therefore do not claim the cross-dataset contrast as a demonstration. The two limbs draw on different primary literatures and cannot be paired plot by plot, and a between-dataset comparison of this size is underpowered by construction. The demonstration rests instead on the two tests that are internal to a single dataset: the within-study divergence above, and the turnover gradient in Section 3.7.

## 3.5 The pattern is not confined to plant invaders

Invasive rats on 18 New Zealand offshore islands, which suppress seabird colonies and with them the marine nutrient subsidy, were associated with live plant biomass carbon 104% higher and non-living carbon pools 26% lower, for a net increase in total ecosystem carbon of 37% [@wardle2007promotion]. Three features of this result matter here. The two pools move in opposite directions, so a study reporting either alone would reach a different conclusion from one reporting both. The effect is entirely indirect, operating through predation on seabirds and the consequent loss of a marine nutrient subsidy, with no plant invasion involved. And the invader is a predator, so the mechanism cannot be attributed to the growth traits usually invoked to explain invader carbon effects.

Carbon effects of invasion therefore arise across trophic levels, consistent with the framework of @peltzer2010effects, who identify characteristic direct and indirect pathways for primary producers, decomposers, herbivores, pathogens, mutualists and predators alike. The compiled evidence is nevertheless dominated by primary producers, and the single predator case, although well designed, cannot establish generality on its own.

## 3.6 Heterogeneity, small-study effects and sensitivity

```{r}
#| label: tbl-sens
#| echo: true
#| tbl-cap: "Sensitivity of the pooled soil organic carbon estimate to model specification and to the treatment of observations with imputed dispersion."
sub <- soc[!soc$sd_imputed, ]
m_rep <- rma.mv(yi, vi, random = ~ 1 | study/obs, data = sub, method = "REML")
m_sl  <- rma(yi, vi, data = soc, method = "REML")
m_egg <- rma.mv(yi, vi, mods = ~ sqrt(vi), random = ~ 1 | study/obs, data = soc, method = "REML")
t3 <- data.frame(
  Specification = c("Multilevel, imputation applied (primary)", "Multilevel, reported dispersion only",
                    "Multilevel, observations weighted equally", "Single-level random effects"),
  `Change (%)` = c(sg(r_soc$est), sg(pct(as.numeric(m_rep$b))),
                   sg(pct(as.numeric(m_unw$b))), sg(pct(as.numeric(m_sl$b)))),
  `95% CI` = c(sprintf("%s to %s", sg(r_soc$lo), sg(r_soc$hi)),
               sprintf("%s to %s", sg(pct(m_rep$ci.lb)), sg(pct(m_rep$ci.ub))),
               sprintf("%s to %s", sg(pct(m_unw$ci.lb)), sg(pct(m_unw$ci.ub))),
               sprintf("%s to %s", sg(pct(m_sl$ci.lb)), sg(pct(m_sl$ci.ub)))),
  Observations = c(r_soc$k, m_rep$k, m_unw$k, m_sl$k), check.names = FALSE)
kable(t3, row.names = FALSE)
write.csv(t3, file.path(repo, "03.outputs/tables/table-sensitivity.csv"), row.names = FALSE)
```

Heterogeneity was at least `r f1(min(res$I2))`% for every fraction, which is high even by the standards of ecological synthesis [@senior2016heterogeneity] and means the pooled estimates describe a distribution of effects rather than a common one. There was no evidence of small-study effects in the soil organic carbon subset (regression of effect size on standard error, *b* = `r f1(m_egg$b[2])`, *p* `r pf(m_egg$pval[2])`). Dispersion was imputed for only `r f1(100*mean(d$sd_imputed))`% of observations, and excluding them changed the pooled estimate by less than one percentage point. The estimate was sensitive to weighting rather than to imputation, and in no specification did the interval exclude zero.

## 3.7 The effect declines with the turnover time of the pool

```{r}
#| label: fit-turnover
#| echo: true

# Turnover classes are assigned from each fraction's published characterisation, before any
# model is fitted and without reference to the results. Bulk measures aggregate all classes and
# are excluded from the ordered test.
TURNOVER <- c(MBC = "fast", DOC = "fast", WSOC = "fast", ROC = "fast",
              POC = "intermediate", MAOC = "slow", SOC = "mixed", TC = "mixed")
RANK <- c(fast = 1, intermediate = 2, slow = 3)
d$turnover  <- unname(TURNOVER[d$fraction])
d$turn_rank <- unname(RANK[d$turnover])

tcls <- do.call(rbind, lapply(c("fast", "intermediate", "slow"), function(k) {
  f <- fit(d[which(d$turnover == k), ]); if (is.null(f)) return(NULL)
  data.frame(turnover = k, k_obs = f$k, k_study = f$ns, pct = f$est,
             lo = f$lo, hi = f$hi, p = f$p, stringsAsFactors = FALSE)}))
write.csv(tcls, file.path(TBL, "table-turnover-classes.csv"), row.names = FALSE)

# Rank enters as a continuous moderator, which tests the ordered prediction directly rather
# than testing for any difference among classes; the categorical form confirms the result is
# not an artefact of the linear coding.
sub_t <- d[!is.na(d$turn_rank), ]
tord <- rma.mv(yi, vi, mods = ~ turn_rank, random = ~ 1 | study/obs, data = sub_t, method = "REML")
tcat <- rma.mv(yi, vi, mods = ~ factor(turnover), random = ~ 1 | study/obs,
               data = sub_t, method = "REML")

kable(data.frame(`Turnover class` = tcls$turnover, Observations = tcls$k_obs,
                 Studies = tcls$k_study, `Change (%)` = sg(tcls$pct),
                 `95% CI` = sprintf("%s to %s", sg(tcls$lo), sg(tcls$hi)),
                 p = signif(tcls$p, 3), check.names = FALSE), row.names = FALSE)
```

```{r}
#| label: fig-turnover
#| echo: true
#| fig-width: 7.0
#| fig-height: 4.2
#| fig-cap: "Effect of invasion on carbon by the turnover class of the pool, with classes assigned a priori from published characterisations. Points are percentage changes with 95% confidence intervals from multilevel random-effects models; filled symbols indicate intervals excluding zero. Numbers at right give observations and, in parentheses, contributing studies."

TL <- c(fast = "Fast\n(days to months)", intermediate = "Intermediate\n(years to decades)",
        slow = "Slow\n(decades to centuries)")
o5 <- tcls[rev(seq_len(nrow(tcls))), ]
draw_fig5 <- function()
  forest_panel(o5$pct, o5$lo, o5$hi, o5$p, TL[o5$turnover],
               sprintf("k=%d (%d)", o5$k_obs, o5$k_study),
               "Change in carbon under invasion (%)", lmar = 8.0)
png(file.path(FIG, "figure-5-turnover.png"), width = 2100, height = 1250, res = 300)
draw_fig5(); invisible(dev.off())
draw_fig5()
```

The prediction was supported. Entering turnover rank as a continuous moderator gave a slope of `r sprintf("%.3f", tord$b[2])` (95% CI `r sprintf("%.3f", tord$ci.lb[2])` to `r sprintf("%.3f", tord$ci.ub[2])`, *Q*~M~ = `r f1(tord$QM)`, df = `r tord$m`, *p* `r pf(tord$QMp)`), across `r tord$k` observations from `r length(unique(sub_t$study))` studies. Each step toward slower turnover was associated with a change in the invasion effect of `r sg(pct(tord$b[2]))`%. The categorical specification agreed (*Q*~M~ = `r f1(tcat$QM)`, df = `r tcat$m`, *p* `r pf(tcat$QMp)`).

The class estimates were `r sg(tcls$pct[tcls$turnover=="fast"])`% for fast pools (95% CI `r sg(tcls$lo[tcls$turnover=="fast"])` to `r sg(tcls$hi[tcls$turnover=="fast"])`, `r tcls$k_obs[tcls$turnover=="fast"]` observations from `r tcls$k_study[tcls$turnover=="fast"]` studies), `r sg(tcls$pct[tcls$turnover=="intermediate"])`% for intermediate pools (95% CI `r sg(tcls$lo[tcls$turnover=="intermediate"])` to `r sg(tcls$hi[tcls$turnover=="intermediate"])`), and `r sg(tcls$pct[tcls$turnover=="slow"])`% for slow pools (95% CI `r sg(tcls$lo[tcls$turnover=="slow"])` to `r sg(tcls$hi[tcls$turnover=="slow"])`).

Two qualifications are necessary. The slow class rests on only `r tcls$k_obs[tcls$turnover=="slow"]` observations from `r tcls$k_study[tcls$turnover=="slow"]` studies, its interval is wide, and its point estimate lies above rather than below the intermediate class, so it does not sit on the fitted line. The ordered result is therefore driven principally by the contrast between fast and intermediate pools, which is well supported, rather than by a demonstrated monotonic decline across all three. Resolving the slow class would require substantially more paired measurements of mineral-associated carbon, and that measurement is the single most valuable addition the field could make to this question.

```{r}
#| label: abstract-guard
#| echo: true

# The abstract sits in the YAML front matter, which Quarto parses before any chunk runs, so it
# cannot carry live inline expressions and its figures must be literals. This guard recomputes
# every one of them and stops the render if the abstract has drifted from the analysis, which
# preserves the rule that no number in the manuscript can disagree with the code. It sits at the
# end of the Results because it depends on every model fitted above.
ABSTRACT_CLAIMS <- list(
  n_obs        = list(text = 425,   value = nrow(d)),
  n_studies    = list(text = 57,    value = length(unique(d$study))),
  soc_all      = list(text = 20.4,  value = r_soc$est),
  soc_unw      = list(text = 6.8,   value = pct(as.numeric(m_unw$b))),
  soc_areal    = list(text = -0.8,  value = r_area$est),
  soc_areal_lo = list(text = -15.7, value = r_area$lo),
  soc_areal_hi = list(text = 16.7,  value = r_area$hi),
  n_flux       = list(text = 688,   value = n_flux),
  n_site       = list(text = 110,   value = n_site),
  n_country    = list(text = 25,    value = n_ctry),
  resp         = list(text = 24.3,  value = q_resp$pct),
  cenz         = list(text = 24.6,  value = q_cenz$pct),
  pool         = list(text = 12.6,  value = q_pool$pct),
  fauna        = list(text = -32.3, value = q_faun$pct),
  contrast_gap = list(text = 25.3,  value = ca_gap),
  contrast_p   = list(text = 0.156, value = round(ca$QMp, 3)))
drift <- vapply(ABSTRACT_CLAIMS, function(v) abs(v$text - v$value) > 0.05, TRUE)
if (any(drift))
  stop("Abstract has drifted from the analysis for: ",
       paste(names(ABSTRACT_CLAIMS)[drift], collapse = ", "),
       ". Update the YAML abstract to the recomputed values.")
cat("abstract verified against the analysis:", length(ABSTRACT_CLAIMS), "figures, no drift\n")
```

# Discussion

## 4.1 Rate and stock are not interchangeable

Invasion did not detectably increase soil carbon stocks in this synthesis, while the processes that move carbon through the soil ran measurably faster: respiration, enzyme activity and microbial biomass were all elevated. The two findings are not in conflict once it is recognised that they concern different quantities. Stock is the integral of rate over residence time, and a system can accumulate carbon quickly while retaining little.

The mechanism was set out by @peltzer2010effects, who concluded that direct invader effects on forest carbon are smaller and shorter-lived than indirect effects operating through nutrient availability, productivity and species composition, and that these indirect effects govern long-term pools. Rate is measurable within a season and responds immediately to a fast-growing invader. Stock integrates decades of gains and losses and responds only if residence time is preserved. Our fraction results are consistent with residence time falling rather than rising: microbial biomass carbon increased while particulate organic carbon, composed largely of partially decomposed plant residues, decreased, a signature of faster processing rather than accrual. The flux limb supports the same reading from the process side: soil respiration and carbon-acquiring enzyme activity both rose by roughly a quarter, consistent with the reported increase in heterotrophic respiration under invasion [@fan2024global]. One published expectation was not met. @zhou2019enhanced report that invasion raises nutrient-releasing enzyme activity while leaving carbon-decomposing enzymes inconsistent, which predicts an asymmetry between the two enzyme groups; we found none (*p* `r pf(m_enz$QMp)`), with both groups elevated by a similar margin. On these data the acceleration is general rather than specific to carbon acquisition, and the mechanism is better described as intensified microbial activity overall than as a selective release of carbon-degrading capacity.

The turnover test converts this interpretation from a reading of two fractions into a quantitative prediction that the data confirm. Ordering pools by their published turnover time and entering that order as a moderator, the invasion effect declines by `r f1(abs(pct(tord$b[2])))`% per step from fast to slow (*p* `r pf(tord$QMp)`). Invasion therefore does not add carbon uniformly across the soil profile's kinetic classes: it adds carbon to the compartments that turn over fastest and does not add it, and may remove it, from those that persist. Since only the persistent compartments contribute to storage on the timescales relevant to climate mitigation, an increase concentrated in fast pools is consistent with accelerated throughput and inconsistent with accrual. The weakest point in this test is the slow class itself, which is too sparsely measured to resolve, so the inference rests on the fast-to-intermediate contrast rather than on the full ordering.

## 4.2 Two measurement artefacts inflate the apparent stock effect

Two features of how stocks are measured act in the same direction as the rate-stock confusion.

The apparent soil carbon gain existed only in concentration data and vanished among areal stocks. Concentration and stock differ by bulk density, which invasion alters in both directions: it is reduced beneath encroaching woody legumes, where ignoring the change overestimates subcanopy soil carbon by nearly thirty per cent [@throop2008shrub], and the *Spartina* literature notes explicitly that bulk density is often reduced under invasion [@xu2022effects]. The IPCC states the consequence directly, that a fixed-depth stock change can arise entirely from a bulk density change [@ipcc2019refinement], while confining the remedy to higher-tier factor derivation; the 2006 Guidelines, still operative for most inventories, treat bulk density only as a multiplier [@ipcc2006guidelines]. The correction, available for three decades [@ellert1995calculation] and repeatedly shown to matter [@raffeld2024importance; @potash2026empirical], cannot be applied here because no contributing study reported bulk density. That absence is a principal finding: the primary literature is not currently constituted to answer whether invasion changes soil carbon stocks.

Second, the positive response was confined to coastal systems and dominated by one species. This is a rediscovered rather than a novel pattern. The largest published positive result was attributed by its own authors to invaders that "intensify biomass per unit area, extend and elevate coastal wetlands, and convert coastal mudflats into C-rich vegetated habitat" [@davidson2018differential]; where the comparison is against bare tidal flat the difference is very large [@huang2022effects], and where it is against established native marsh it is smaller and expressed as concentration [@wang2018spartina; @xu2022effects]. Sampling conventions compound this: the generic IPCC soil depth default is 30 cm while coastal wetlands are assessed to 1 m [@ipcc2006guidelines; @ipcc2014wetlands], so the systems carrying the positive effect are those conventionally sampled deepest.

## 4.3 The accounting consequence

The IPCC treats the gain-loss and stock-difference methods as equally valid and leaves the choice to expert judgement [@ipcc2006guidelines]. On invaded land that choice determines the answer. A gain-loss assessment, built from increment and loss terms, will register the elevated productivity that invasion reliably produces. A stock-difference assessment, built from inventories separated in time, will register little or nothing, and the guidance itself anticipates this, warning that stock-difference fails where change is small relative to the standing stock.

This matters because carbon crediting pays for durable stock while frequently being justified by rate-like evidence. Projections built on observed growth rates rather than realised stock differences will overstate the credited quantity wherever residence time is shorter than assumed. The direction of error in the land sector is already documented: improved-forest-management baselines have been shown to understate natural accumulation, inflating the credited difference [@west2023action], and an assessment of 182 Australian regeneration projects found changes in credited areas largely mirrored changes in adjacent uncredited land [@macintosh2024australian]. Invasion is a case where rate and stock evidence are both available and disagree, and where the disagreement is mechanistic rather than incidental.

It is worth recording how little accounting frameworks engage with invasion. IPCC carbon pools are defined as physical compartments distinguished by whether material is living or dead, above or below ground, and above or below a size threshold, with no reference to species identity or provenance [@ipcc2006guidelines]. Invasion is consequently invisible to national accounts: an invaded stand and a native stand of equal biomass are indistinguishable, and the guidance supplies default biomass values for species that are aggressive invaders elsewhere. The one peer-reviewed treatment of this gap estimates that invasions of ten problem plants may emit up to 150 Mt CO~2~-equivalent annually in a single country [@mantoani2026invasive].

## 4.4 Bio-perversity

The rate-stock divergence bears directly on a recognised policy risk. @lindenmayer2012avoiding identify three bio-perversities arising from carbon programmes, the second being the planting of trees that become invasive taxa. If carbon programmes reward measured accumulation rates, they will preferentially select fast-growing species, and fast growth is among the traits that predict invasiveness. @pati2025ecological, having found rapid carbon accumulation under *Gliricidia sepium* alongside reduced diversity, conclude that planting invasive species "might create obstruction to achieve the goals of REDD+ and UN resolutions on ecosystem restoration", and title their paper an ecological dilemma.

The contrary case has been argued: that invader biomass should be valorised through established value chains [@nunes2021carbon], or that in degraded systems an invader may build soil carbon more effectively than natives through root-derived inputs to mineral-associated pools [@meitha2024invasive]. Our results do not refute these arguments, which concern particular species in particular settings, but they do bear on the general form of the claim. Where the evidence offered is a rate, it is not evidence about stock, and it is stock that a carbon account records.

## 4.5 What would follow if the divergence is real

Three consequences follow, in ascending order of importance.

For measurement, rate and stock should be reported together. They are currently reported by different subfields using different designs, and a study measuring one has no way of knowing whether the other agrees. @pati2025ecological demonstrate what becomes visible when both are measured on the same plots, and theirs is a rare design rather than a standard one.

For interpretation, evidence of accelerated accumulation should not be presented as evidence of increased storage. This is not a pedantic distinction. In the compiled evidence the two differ by nearly an order of magnitude in the same system, and they differ in sign between compartments in the same study. The literature's apparent disagreement about invasion and carbon substantially dissolves once the quantity being reported is specified.

For policy, the mismatch is systematic rather than incidental. A crediting system that accepts rate-like evidence while issuing credits against stock will over-credit wherever residence time is shorter than assumed, and invasion is a case where there is positive reason to expect exactly that, because the fraction data indicate faster cycling. The direction of this error is the same as that already documented in improved forest management, where baselines understate natural accumulation [@west2023action], and in regeneration projects whose credited areas track adjacent uncredited land [@macintosh2024australian].

The counter-arguments deserve a fair statement. @nunes2021carbon propose valorising invader biomass through established value chains, which is a proposal about the fate of harvested material rather than about standing stock, and is unaffected by our result: material converted to a durable product leaves the land pool and is accounted elsewhere. @meitha2024invasive argue that root-derived carbon from a deep-rooted invader may form mineral-associated soil carbon more efficiently than litter inputs, which is a claim specifically about the durable fraction and is therefore the right kind of claim. Our data cannot test it, because mineral-associated carbon was represented by only `r r_maoc$k_obs` observations from `r r_maoc$k_study` studies. That is the measurement most needed, and its absence is the reason the durability question remains open rather than closed.

## 4.6 Limitations

The central weakness is that the two limbs cannot be paired. Stocks and fluxes are each estimated from a formal meta-analysis, but from different deposits covering different primary literatures, so no site contributes both quantities and the contrast between them is necessarily between datasets. Fitted directly, that contrast was `r sg(ca_gap)` percentage points in the predicted direction but did not reach significance (*p* `r pf(ca$QMp)`), and with `r ca$k - r_area$k` flux observations against `r r_area$k` areal stock observations it was not well placed to. We report it as a consistency check that the argument passes rather than as a test it survives, and the demonstration rests on the two within-dataset results instead. What the field needs is not a larger synthesis of either quantity separately but studies that measure both on the same plots, of which @pati2025ecological is a rare example.

The flux limb carries its own limitations. Its deposit records no complete study identifier, so observations were clustered by site; where one publication contributed several sites this treats them as independent and the intervals are correspondingly narrow. Soil respiration, the only direct carbon flux available, rested on `r q_resp$k_obs` observations from `r q_resp$k_study` sites, and within forests alone on `r m_frs$k` observations from `r n_forest_resp_site` sites, too few to resolve. The result is therefore cross-biome and is not demonstrated for forests specifically, which matters because forest was also the habitat class showing no stock response. Enzyme activities and microbial pools are not fluxes; they are read here as evidence on mechanism, and a change in potential enzyme activity assayed under standard conditions need not imply a proportional change in realised decomposition in the field.

The rate compilation is weaker still, drawing on four sources, tabulated rather than pooled, and conclusions from it are drawn as such.

The stock analysis rests on a single deposited compilation and inherits its inclusion criteria and geographical emphasis. Heterogeneity was at least `r f1(min(res$I2))`% for every fraction, so pooled estimates describe a distribution rather than a common effect [@senior2016heterogeneity]. The areal subset carrying the central result comprises `r r_area$k` observations from `r n_area_stud` studies. Mineral-associated carbon was too sparsely represented to test. All comparisons are spatial contrasts without pre-invasion measurement, so pre-existing differences influencing where invaders established cannot be excluded, and mechanisms underlying reported impacts are rarely tested directly in this literature [@levine2003mechanisms].

# Conclusions

Biological invasion increases the rate at which carbon moves through an ecosystem without reliably increasing the stock that is held. Respiration, enzyme activity and microbial biomass all rose; areal soil carbon did not. The distinction is not semantic: it corresponds to the two methods the IPCC treats as equally valid, it is visible within single datasets, and it extends beyond plant invaders to effects transmitted through trophic cascades. It is also not yet demonstrated by a direct test, because no deposited dataset measures both quantities on the same plots, and closing that gap is the single most useful thing the field could do with this question.

Three changes would resolve what remains uncertain. Studies of invasion and carbon should report rates and stocks together, since the divergence is only detectable when both are measured. Soil studies should report bulk density alongside concentration, sample below 30 cm, and express stocks on an equivalent soil mass basis. And carbon accounting for invaded or restored land should state which of the two IPCC methods it applies, because on this land the choice determines the result. Until then, evidence that invasion accelerates carbon accumulation should not be read as evidence that it stores more carbon.

## Data availability

Both source datasets are openly available from Dryad: the stock data at doi:10.5061/dryad.c59zw3rkc [@he2025plant] and the flux data at doi:10.5061/dryad.hhmgqnkq5 [@negesse2025invasive]. The compiled rate evidence, the code that tidies both deposits, fits the models and produces every figure and table, and the manuscript source are archived together and require only R and the `metafor` package.

## References

::: {#refs}
:::
